---
title: "Do NBA players get tired? A case-study of Giannis Antetokounmpo"
description: |
  NBA players are ATHLETES. But, do they get tired? Does their performance decline late in the game? 
author:
  - name: Guy Frederick Sutton
    url: {}
date: 01-25-2021
output:
  distill::distill_article:
    self_contained: false
draft: false
base_url: https://nbaanalyticsR.netlify.app/
categories:
  - R
  - tidyverse
  - Data cleaning
  - ggplot2
  - Player performance
  - Milwaukee Bucks
  - Giannis Antetokounmpo
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

It always amazes me watching professional athletes at work. If I had to run, shoot, defend, sprint, screen like the NBA pros do... I would be lying on the floor, gasping for air within the 1st quarter. This got me thinking, do NBA players get tired? Does their performance vary over time? Do they make fewer rebounds or field goals the longer they have been on the court? While this is certainly a topic that peeks my interest, there are also some practical implications. For example, it could be quite informative to identify optical performance thresholds for each player to know:

- (1) how long can they perform at their peek abilities per game? (e.g. how many minutes per game is optimal for their performance?), and 
- (2) how long each stint on court during the game should be to maximise performance? 

In my first blog post, I will be exploring these two questions using some basic data visualisations. Later blogposts will get into a bit more detail (building some statistical models) - stay tuned! 

# Load required packages 

The workhorse package we will be using is the NBA stats API package [`nbastatR`](https://github.com/abresler/nbastatR). We need to install the package from GitHub, which only needs to be done once. 

```{r install git package, include = TRUE, echo = TRUE, warning = FALSE, message = FALSE, results = "hide"}
# To install a GitHub package: 
devtools::install_github("abresler/nbastatR")

# Once you have downloaded a GitHub package, we should then
# edit our code so that the package is not re-installed each time we 
# re-run our script. We place a '#' in front of the command to comment it out. 
# If we need to run that code again later, we simply remove the '#' 
# devtools::install_github("abresler/nbastatR")
```

Our next step is to load the other packages we will require for this analysis. I really like using the `pacman` package to install/load R packages. I find this really helps when sharing code with collaborators who are less familiar with R. They can just open the script and press 'run' and not have to worry about installing new packages. `pacman` will install any packages that are not installed on your machine, and load all the remaining packages. 

```{r, include = TRUE, echo = TRUE, results = "hide"}
# Load required packages 
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               lubridate,
               zoo,
               janitor,
               nbastatR,
               future,
               scales)
```

```{r set ggplot theme, include = FALSE, echo = TRUE, results = "hide"}
# Set ggplot theme (makes nice plots)
theme_set(theme_classic() +
            theme(panel.border = element_rect(colour = "black", fill = NA),
                  axis.text = element_text(colour = "black"),
                  axis.title.x = element_text(margin = unit(c(2, 0, 0, 0), "mm")),
                  axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm")),
                  legend.position = "none"))
```

# Downloading play-by-play data 

We will now use the `nbastatR` package to download play-by-play data from the NBA API. We can download data for different seasons, different teams... This can take a really long time (at least it does on my PC), so below, I will just download the data for (1) the 2019/2020 season, and (2) only the Milwaukee Bucks. 


```{r download game logs, echo = TRUE, results = "hide", include = TRUE, warnings = FALSE, messages = FALSE, cache = TRUE}
# Extract all games for the 2020 season (excluding pre-season and all-star weekend)
game_logs <- game_logs(seasons = 2020)

# Process the game logs 
games <- game_logs %>%
  # Keep only these columns 
  dplyr::select(idGame, slugTeam, slugOpponent, locationGame) %>%
  # Use a conditional statement to process the team home/away data 
  dplyr::mutate(slugTeamHome = dplyr::if_else(locationGame == "H", 
                                              slugTeam, 
                                              slugOpponent),
                slugTeamAway = dplyr::if_else(locationGame == "A", 
                                              slugTeam, 
                                              slugOpponent)) %>%
  # Drop these columns - no longer required
  dplyr::select(-c(slugTeam, slugOpponent, locationGame)) %>%
  dplyr::distinct(idGame, .keep_all = TRUE) %>%
  # Here, we filter for the teams we want data for 
  # We have to use team codes. E.g. MIL = Milwaukee Bucks
  # - Filter the games where either (denoted by the '|' bar) the home or 
  #   or away team is the Bucks. 
  dplyr::filter(slugTeamHome == "MIL" | slugTeamAway == "MIL")

# Extract the play-by-play for each game 
play_logs_all <- play_by_play_v2(game_ids = unique(games$idGame))
```

Once the data is downloaded, quite a bit of data pre-processing and data cleaning is required, which will not be shown in this blogpost. I have edited this code slightly from the amazing [NBA blog by Ramiro Bentes](https://nbainrstats.netlify.app/post/adding-lineups-to-nba-play-by-play-data/). If you want to delve into the data cleaning code, it is available in the [source R markdown file] (https://github.com/guysutton/nba_analytics_blog/blob/main/_posts/2021-01-25-do-nba-players-get-tired/do-nba-players-get-tired.Rmd) for this blogpost. 

```{r background data cleaning, include = FALSE, results = "hide"}
############################################################
# - Pre-processing data cleaning
############################################################

# Lots, and lots, and lots of data cleaning 
new_pbp <- play_logs_all %>%
  dplyr::distinct(idGame, numberEvent, .keep_all = TRUE) %>%
  dplyr::group_by(idGame) %>%
  dplyr::mutate(numberEvent = row_number()) %>%  
  dplyr::ungroup() %>%
  dplyr::select(idGame, numberEventMessageType, numberEventActionType, 
                namePlayer1, namePlayer2, namePlayer3,  
                slugTeamPlayer1, slugTeamPlayer2,  slugTeamPlayer3, 
                numberPeriod, timeQuarter, minuteRemainingQuarter,  
                secondsRemainingQuarter, descriptionPlayHome, numberEvent, 
                descriptionPlayVisitor, scoreHome, scoreAway) %>%
  dplyr::mutate(shotPtsHome = dplyr::case_when(
    numberEventMessageType == 3 & !stringr::str_detect(descriptionPlayHome, "MISS") ~ 1,         
    numberEventMessageType == 1 & stringr::str_detect(descriptionPlayHome, "3PT") ~ 3,          
    numberEventMessageType == 1 & !stringr::str_detect(descriptionPlayHome, "3PT") ~ 2,
    TRUE ~ 0)) %>%
  dplyr::mutate(shotPtsAway = dplyr::case_when(
    numberEventMessageType == 3 & !stringr::str_detect(descriptionPlayVisitor, "MISS") ~ 1,
    numberEventMessageType == 1 & stringr::str_detect(descriptionPlayVisitor, "3PT") ~ 3,
    numberEventMessageType == 1 & !stringr::str_detect(descriptionPlayVisitor, "3PT") ~ 2,
    TRUE ~ 0)) %>%
  dplyr::group_by(idGame) %>%
  dplyr::mutate(ptsHome = cumsum(shotPtsHome),
                ptsAway = cumsum(shotPtsAway)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(secsLeftQuarter = (minuteRemainingQuarter * 60) + secondsRemainingQuarter) %>%
  dplyr::mutate(secsStartQuarter = dplyr::case_when(
    numberPeriod %in% c(1:5) ~ (numberPeriod - 1) * 720,
    TRUE ~ 2880 + (numberPeriod - 5) * 300)) %>%
  dplyr::mutate(secsPassedQuarter = ifelse(numberPeriod %in% c(1:4), 
                                           720 - secsLeftQuarter, 
                                           300 - secsLeftQuarter),
                secsPassedGame = secsPassedQuarter + secsStartQuarter) %>%
  dplyr::left_join(games %>%
                     dplyr::select(idGame, slugTeamHome, slugTeamAway)) %>%
  dplyr::select(idGame, numberEventMessageType, numberEventActionType, 
                slugTeamHome, slugTeamAway, slugTeamPlayer1, slugTeamPlayer2, 
                slugTeamPlayer3, numberPeriod, timeQuarter, secsPassedGame, numberEvent, 
                namePlayer1, namePlayer2, namePlayer3, descriptionPlayHome,
                descriptionPlayVisitor, ptsHome, ptsAway, shotPtsHome, shotPtsAway) %>%
  dplyr::mutate(marginBeforeHome = ptsHome - ptsAway - shotPtsHome + shotPtsAway,
                marginBeforeAway = ptsAway - ptsHome - shotPtsAway + shotPtsHome,
                timeQuarter = stringr::str_pad(timeQuarter, width = 5, pad = 0))

# Define where substitutes where made 
subs_made <- new_pbp %>%
  dplyr::filter(numberEventMessageType == 8) %>%
  dplyr::mutate(slugTeamLocation = ifelse(slugTeamPlayer1 == slugTeamHome, "Home", "Away")) %>%
  dplyr::select(idGame, numberPeriod, timeQuarter, secsPassedGame, 
                slugTeamPlayer = slugTeamPlayer1,
                slugTeamLocation, 
                playerOut = namePlayer1, 
                playerIn = namePlayer2) %>%
  tidyr::pivot_longer(cols = starts_with("player"), 
                      names_to = "inOut",
                      names_prefix = "player",
                      values_to = "namePlayer") %>%
  dplyr::group_by(idGame, numberPeriod, slugTeamPlayer, namePlayer) %>%
  dplyr::filter(row_number() == 1) %>%
  dplyr::ungroup()

others_qtr <- new_pbp %>%
  dplyr::filter(numberEventMessageType != 8) %>%                             
  dplyr::filter(!(numberEventMessageType == 6 & 
                    numberEventActionType %in% c(10, 11, 16, 18, 25))) %>%
  tidyr::pivot_longer(cols = starts_with("namePlayer"),
                      names_to = "playerNumber",
                      names_prefix = "namePlayer",
                      values_to = "namePlayer") %>%
  dplyr::mutate(slugTeamPlayer = dplyr::case_when(playerNumber == 1 ~ slugTeamPlayer1,
                                                  playerNumber == 2 ~ slugTeamPlayer2,
                                                  playerNumber == 3 ~ slugTeamPlayer3,
                                                  TRUE ~ "None")) %>%
  dplyr::mutate(slugTeamLocation = ifelse(slugTeamPlayer == slugTeamHome, 
                                          "Home", 
                                          "Away")) %>%
  dplyr::filter(!is.na(namePlayer),
                !is.na(slugTeamPlayer)) %>%
  dplyr::anti_join(subs_made %>%
                     dplyr::select(idGame, numberPeriod, slugTeamPlayer, namePlayer)) %>%  
  dplyr::distinct(idGame, numberPeriod, namePlayer, slugTeamPlayer, slugTeamLocation)

lineups_quarters <- subs_made %>%
  dplyr::filter(inOut == "Out") %>%
  dplyr::select(idGame, numberPeriod, slugTeamPlayer, namePlayer, slugTeamLocation) %>%
  dplyr::bind_rows(others_qtr) %>%
  dplyr::arrange(idGame, numberPeriod, slugTeamPlayer)

lineup_subs <- new_pbp %>%
  dplyr::filter(numberEventMessageType == 8) %>%
  dplyr::select(idGame, numberPeriod, timeQuarter, secsPassedGame, 
                slugTeamPlayer = slugTeamPlayer1, 
                playerOut = namePlayer1, 
         playerIn = namePlayer2, numberEvent) %>%
  dplyr::arrange(idGame, numberEvent) %>%
  dplyr::group_by(idGame, numberPeriod, slugTeamPlayer) %>%
  dplyr::mutate(row1 = row_number()) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(lineups_quarters %>%
                     dplyr::group_by(idGame, numberPeriod, slugTeamPlayer) %>%
                     dplyr::summarise(lineupBefore = paste(sort(unique(namePlayer)), 
                                                           collapse = ", ")) %>%
                     dplyr::ungroup() %>%
                     dplyr::mutate(row1 = 1)) %>%
  dplyr::select(-row1)

lineup_subs <- lineup_subs %>%
  dplyr::mutate(lineupBefore = stringr::str_split(lineupBefore, ", ")) %>% 
  dplyr::arrange(idGame, numberEvent) %>%
  dplyr::group_by(idGame, numberPeriod, slugTeamPlayer) %>%
  dplyr::mutate(lineupAfter = accumulate2(playerIn, 
                                          playerOut, 
                                          ~setdiff(c(..1, ..2), ..3), 
                                          .init = lineupBefore[[1]])[-1],
         lineupBefore = dplyr::coalesce(lineupBefore, lag(lineupAfter))) %>%
  dplyr::ungroup() %>% 
  dplyr::mutate_all( ~ purrr::map_chr(., ~paste(.x, collapse = ", "))) %>%
  dplyr::mutate_at(vars("numberEvent", "numberPeriod", "idGame"), 
                   ~ as.integer(.)) %>%
  dplyr::mutate(secsPassedGame = as.numeric(secsPassedGame)) %>%
  dplyr::arrange(idGame, numberEvent) %>%
  dplyr::left_join(lineups_quarters %>%
                     dplyr::distinct(idGame, slugTeamPlayer, slugTeamLocation)) %>%
  dplyr::filter(!is.na(slugTeamLocation))

lineup_game <- new_pbp %>%
  dplyr::group_by(idGame, numberPeriod) %>%
  dplyr::mutate(row1 = row_number()) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(lineups_quarters %>%
                     dplyr::group_by(idGame, numberPeriod, slugTeamLocation) %>%
                     dplyr::summarise(lineupBefore = paste(sort(unique(namePlayer)), 
                                                           collapse = ", ")) %>%
                     dplyr::ungroup() %>%
                     tidyr::pivot_wider(names_from = slugTeamLocation,
                                        names_prefix = "lineupInitial",
                                        values_from = lineupBefore) %>%
                     dplyr::mutate(row1 = 1)) %>%
  dplyr::select(-row1) %>%
  dplyr::left_join(lineup_subs %>%
              mutate(lineupBeforeHome = ifelse(slugTeamLocation == "Home", lineupBefore, NA),
                     lineupAfterHome = ifelse(slugTeamLocation == "Home", lineupAfter, NA),
                     lineupBeforeAway = ifelse(slugTeamLocation == "Away", lineupBefore, NA),
                     lineupAfterAway = ifelse(slugTeamLocation == "Away", lineupAfter, NA)) %>%
              select(idGame, numberPeriod, timeQuarter, secsPassedGame, 
                     numberEvent, slugTeamPlayer1 = slugTeamPlayer,
                     contains("Home"), 
                     contains("Away"))) %>%
  dplyr::mutate_at(vars(c(lineupBeforeHome, lineupAfterHome)), 
                   ~ ifelse(!is.na(lineupInitialHome), 
                            lineupInitialHome, 
                            .)) %>%
  dplyr::mutate_at(vars(c(lineupBeforeAway, lineupAfterAway)), 
                   ~ ifelse(!is.na(lineupInitialAway), 
                            lineupInitialAway, 
                            .)) %>%
  dplyr::select(-starts_with("lineupInitial"))

lineup_game <- lineup_game %>%
  dplyr::group_by(idGame, numberPeriod) %>%
  dplyr::mutate(lineupHome = na.locf(lineupAfterHome, na.rm = FALSE),
                lineupAway = na.locf(lineupAfterAway, na.rm = FALSE),
                lineupHome = ifelse(is.na(lineupHome), 
                                    na.locf(lineupBeforeHome, 
                                            fromLast = TRUE, 
                                            na.rm = FALSE), 
                                    lineupHome),
         lineupAway = ifelse(is.na(lineupAway), 
                             na.locf(lineupBeforeAway, 
                                     fromLast = TRUE, 
                                     na.rm = FALSE), 
                             lineupAway),
         lineupHome = stringr::str_split(lineupHome, ", "),
         lineupAway = stringr::str_split(lineupAway, ", "),
         lineupHome = purrr::map_chr(lineupHome, 
                                     ~ paste(sort(.), 
                                             collapse = ", ")),
         lineupAway = purrr::map_chr(lineupAway, 
                                     ~ paste(sort(.), 
                                             collapse = ", "))) %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(starts_with("lineupBefore"), 
                   starts_with("lineupAfter")))

lineup_game_stats <- lineup_game %>%
  dplyr::mutate(canSub = case_when(numberEventMessageType == 5 & 
                                     !numberEventActionType %in% c(1, 2) ~ 1,
                                   numberEventMessageType == 6 & 
                                     numberEventActionType != 16 ~ 1,
                                   numberEventMessageType == 11 & 
                                     numberEventActionType != 4 ~ 1,
                                   numberEventMessageType == 7 & 
                                     numberEventActionType == 5 ~ 1, 
                                   numberEventMessageType == 4 & 
                                     numberEventActionType == 0 & 
                                     !stringr::str_detect(stringr::str_to_upper(descriptionPlayHome),
                                                   "OFF:") ~ 1,
                                   numberEventMessageType == 4 & 
                                     numberEventActionType == 0 & 
                                     !str_detect(stringr::str_to_upper(descriptionPlayVisitor), 
                                                 "OFF:") ~ 1,
                                   TRUE ~ 0)) %>%
  dplyr::mutate(secsPassedGame2 = ifelse(timeQuarter == "12:00" &
                                           (str_detect(str_to_lower(descriptionPlayHome), 
                                                       "technical") |
                                              str_detect(str_to_lower(descriptionPlayVisitor),
                                                         "technical")), 
                                         secsPassedGame + 0.5, secsPassedGame)) %>% 
  dplyr::group_by(idGame, numberPeriod, secsPassedGame) %>%
  dplyr::mutate(numberNew = ifelse(numberEventMessageType == 3 & 
                                     numberEventActionType == 12, 
                                   paste(numberEvent[numberEventMessageType == 3 &
                                                       numberEventActionType == 11], 
                                         collapse = ", "), 
                                   as.character(numberEvent)),
                numberNew = ifelse(numberEventMessageType == 3 & 
                                     numberEventActionType %in% c(14, 15), 
                                   paste(numberEvent[numberEventMessageType == 3 & 
                                                       numberEventActionType == 13], 
                                         collapse = ", "),
                                   numberNew)) %>%
  dplyr::mutate(numberNew = stringr::str_split(numberNew, ", "),
                numberNew = purrr::map(numberNew, ~as.numeric(.)),
                numberNew = purrr::map2_dbl(numberNew, 
                                            numberEvent, 
                                            ~ max(.x[.x <= .y]))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(idGame, numberNew, numberEvent) %>%
  dplyr::group_by(idGame) %>%
  dplyr::mutate(newptsHome = cumsum(shotPtsHome),
                newptsAway = cumsum(shotPtsAway)) %>%
  dplyr::group_by(idGame, numberPeriod, secsPassedGame2) %>%
  dplyr::mutate(subOpp = cumsum(canSub)) %>%
  dplyr::group_by(idGame = as.character(idGame),
                  numberPeriod = as.character(numberPeriod), 
                  subOpp, 
                  secsPassedGame2 = as.character(secsPassedGame2)) %>%
  dplyr::mutate(hasFouls = sum(numberEventMessageType == 3)) %>%
  dplyr::mutate(newptsHome = ifelse(hasFouls > 0,
                                    newptsHome[row_number() == 
                                                 max(row_number()[numberEventMessageType == 3])],
                                    newptsHome),
                newptsAway = ifelse(hasFouls > 0,
                                    newptsAway[row_number() == 
                                                 max(row_number()[numberEventMessageType == 3])],
                                    newptsAway)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-hasFouls) %>%
  dplyr::select(-c(numberNew, secsPassedGame2)) %>%
  dplyr::mutate_all(~ as.character(.)) %>%
  dplyr::mutate(secsPassedGame = as.numeric(secsPassedGame),
                numberEvent = as.numeric(numberEvent))
```

We now have play-by-play data for all the Milwaukee Bucks games for the 2019/2020 season. Each game has a unique ID number (`idGame`), each play within each game has its own unique ID (`numberEvent`), and for each play, we have the players who were on the court for each team (`homeTeamLineup` and `awayTeamLineup`), amongst a range of other variables.  

```{r, comment = NA}
head(lineup_game_stats)
```

However, we still need to process the data so that we can:

- Define how many periods each player was on the court for during each game
- Define how long was each period 
- Convert our data into long-format, so that each player who was on the court, for each play, gets their own row. (We will need this format of data when we start summarising how player performance later on). 

```{r, warning = FALSE, message = FALSE}
# For each event, create a distinct row for each player from each team 
game_test <- lineup_game_stats %>%
  # Drop the events that are not plays (e.g. technical fouls)
  tidyr::drop_na(namePlayer1) %>%
  # Give a unique identified to each play per game 
  dplyr::group_by(idGame, numberEvent) %>%
  dplyr::mutate(numberEvent = dplyr::cur_group_id()) %>%
  # Duplicate the full team line-up. We need it for later. 
  dplyr::mutate(homeTeamLineup = lineupHome,
                awayTeamLineup = lineupAway) %>%
  # Combine team line-ups into a single column
  tidyr::unite(homeTeamLineup:awayTeamLineup, col = "lineups", sep = ", ") %>%
  tidyr::separate_rows(., 
                       lineupHome, 
                       lineupAway,
                       sep = ",", 
                       convert = TRUE) %>%
  # Remove whitespace from start of player names, if present
  dplyr::mutate(lineupHome = stringr::str_trim(lineupHome, side = c("left")),
                lineupAway = stringr::str_trim(lineupAway, side = c("left"))) %>%
  # Convert players names into long-format so each player now has their own row 
  # for each event in the game
  tidyr::pivot_longer(
    cols = lineupHome:lineupAway,
    names_to = "teamHomeAway",
    values_to = "playerNames") %>%
  dplyr::ungroup() %>%
  # For each event, is the lineup the same as the previous event? 
  dplyr::group_by(idGame) %>%
  dplyr::mutate(lineupChange = lineups != lag(lineups),
                lineupChange = coalesce(lineupChange, FALSE)) %>%
  dplyr::mutate(lineupStint = cumsum(lineupChange)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(idGame, lineupStint, numberEvent) %>%
  # Has that player been subbed? 
  dplyr::group_by(idGame, playerNames) %>%
  dplyr::arrange(idGame, playerNames, numberEvent) %>%
  # Add column which states whether player was subbed or not? 
  dplyr::mutate(player_subbed = dplyr::case_when(
    lag(numberEvent) != numberEvent - 1 ~ "Yes", 
    TRUE ~ "No"))
```

We now know exactly which plays each player was on the court for, but we need to create a column which explicitly notes whether the player was on the court or not (`stints`), and how long each stint on court lasted. 

```{r, warning = FALSE, message = FALSE, echo = TRUE, results = "hide"}
# Define each stint each player has on court 
define_stint <- game_test %>%
  # Define stints as the cumulative sum of times the player is subbed
  # cumsum starts at 0, so we need to + 1 to each value 
  dplyr::mutate(stints_on_court = cumsum(player_subbed == "Yes"),
                stints_on_court = stints_on_court + 1) %>%
  dplyr::ungroup() %>%
  # Define start and end time of each stint with a min/max statement
  dplyr::group_by(idGame, playerNames, stints_on_court) %>%
  dplyr::mutate(time_at_start = min(secsPassedGame),
                time_at_end   = max(secsPassedGame)) %>%
  # Calculate total seconds of each stint
  dplyr::mutate(time_stint_secs = time_at_end - time_at_start) %>%
  # Convert seconds into minutes and seconds 
  dplyr::mutate(time_stint_mins = time_stint_secs / 60) %>%
  # Calculate time since start of each stint 
  dplyr::mutate(time_since_start_stint = secsPassedGame - time_at_start) %>%
  # Manually define 60 second intervals - 
  # I couldn't force ggplot2::cut_width to start @ 0... Help?  
  dplyr::mutate(minute_cats = dplyr::case_when(
    between(time_since_start_stint, 0, 60) ~ "1",
    between(time_since_start_stint, 61, 120) ~ "2",
    between(time_since_start_stint, 121, 180) ~ "3",
    between(time_since_start_stint, 181, 240) ~ "4",
    between(time_since_start_stint, 241, 300) ~ "5",
    between(time_since_start_stint, 301, 360) ~ "6",
    between(time_since_start_stint, 361, 420) ~ "7",
    between(time_since_start_stint, 421, 480) ~ "8",
    between(time_since_start_stint, 481, 540) ~ "9",
    between(time_since_start_stint, 541, 600) ~ "10",
    between(time_since_start_stint, 601, 660) ~ "11",
    between(time_since_start_stint, 661, 720) ~ "12",
    between(time_since_start_stint, 721, 780) ~ "13",
    between(time_since_start_stint, 781, 840) ~ "14",
    between(time_since_start_stint, 841, 900) ~ "15",
    between(time_since_start_stint, 901, 960) ~ "16",
    between(time_since_start_stint, 961, 1020) ~ "17",
    between(time_since_start_stint, 1021, 1080) ~ "18",
    between(time_since_start_stint, 1081, 1140) ~ "19",
    between(time_since_start_stint, 1141, 1200) ~ "20",
    between(time_since_start_stint, 1221, 1280) ~ "21",
    between(time_since_start_stint, 1281, 1340) ~ "22",
    between(time_since_start_stint, 1341, 1400) ~ "23",
    between(time_since_start_stint, 1401, 1460) ~ "24")) %>%
  # For each play, does the score change or not?
  # - We will need this to classify whether 2pt/3pt/free throw attempts
  #   were successful or not. 
  dplyr::mutate(score_change = dplyr::if_else(shotPtsHome > 0 | 
                                                shotPtsAway > 0, 1, 0))
```

Our last bit of data cleaning requires that we categorize each play. For example, was the play a 2pt field goal attempt, a free throw attempt, was the shot made or missed? 

```{r, warning = FALSE, message = FALSE, echo = TRUE, results = "hide"}
# Combine the descriptions for each play for home and away teams into 
# a single column
description_play <- define_stint %>%
  dplyr::mutate(descriptionPlay = dplyr::coalesce(descriptionPlayHome, descriptionPlayVisitor))

# Define each event as a particular play
events_defined <- description_play %>%
  # Was 3pt fg attempt made?
  dplyr::mutate(is_3pt_made = dplyr::case_when(
    stringr::str_detect(descriptionPlay, "3PT") &
      score_change == 1 ~ 1,
    stringr::str_detect(descriptionPlay, "3PT") &
      score_change == 0 ~ 0)) %>%
  # Was 2pt fg attempt made?
  dplyr::mutate(is_2pt_made = dplyr::case_when(
    numberEventMessageType == 1 & 
      !stringr::str_detect(descriptionPlay, "3PT") ~ 1,
    numberEventMessageType == 2 & 
      !stringr::str_detect(descriptionPlay, "3PT") ~ 0)) %>%
  # Was event a rebound? 
  dplyr::mutate(is_rebound = dplyr::case_when(
    numberEventMessageType == 4 & 
      stringr::str_detect(descriptionPlay, "REBOUND") ~ 1,
    TRUE ~ 0)) %>%
  # Was event a made free throw?
  dplyr::mutate(is_ft_made = dplyr::case_when(
    numberEventMessageType == 3 & 
      score_change == 1 ~ 1,
    numberEventMessageType == 3 & 
      score_change == 0 ~ 0))
```

# Summarise and visualise performance over time 

Finally, the bit of the analysis that we are actually most interested in. Let's now summarise how many 2pt/3pt/free throws/rebounds are made per minute that each player is on the court, for each game, and then these summarise performance metrics over the entire 2020 season. 

```{r, warning = FALSE, message = FALSE, echo = TRUE, results = "hide"}
perform_stint_min <- events_defined %>%
  dplyr::ungroup() %>%
  dplyr::filter(playerNames == "Giannis Antetokounmpo") %>%
  dplyr::filter(namePlayer1 == "Giannis Antetokounmpo") %>%
  dplyr::group_by(playerNames, idGame, stints_on_court, minute_cats, .drop = FALSE) %>%
  dplyr::summarise(
    total_rebounds = sum(is_rebound, na.rm = TRUE),
    total_2pt_made = sum(is_2pt_made, na.rm = TRUE),
    total_2pt_attempt = sum(!is.na(is_2pt_made)),
    perc_2pt_made = total_2pt_made / total_2pt_attempt,
    total_3pt_made = sum(is_3pt_made, na.rm = TRUE),
    total_3pt_attempt = sum(!is.na(is_3pt_made)),
    perc_3pt_made = total_3pt_made / total_3pt_attempt,
    total_ft_made  = sum(is_ft_made, na.rm = TRUE),
    total_ft_attempt = sum(!is.na(is_ft_made)),
    perc_ft_made = total_ft_made / total_ft_attempt) %>%
  dplyr::ungroup()%>%
  # Now calculate mean summary for season (not per-game)
  dplyr::group_by(playerNames, stints_on_court, minute_cats, .drop = FALSE) %>%
  dplyr::summarise(
    n_rebounds       = mean(total_rebounds, na.rm = TRUE),
    n_2pt_made       = mean(total_2pt_made, na.rm = TRUE),
    n_3pt_made       = mean(total_3pt_made, na.rm = TRUE),
    n_ft_attempt     = mean(total_ft_made, na.rm = TRUE),
    sd_rebounds      = sd(total_rebounds, na.rm = TRUE),
    sd_2pt_made      = sd(total_2pt_made, na.rm = TRUE),
    sd_3pt_made      = sd(total_3pt_made, na.rm = TRUE),
    sd_ft_attempt    = sd(total_ft_made, na.rm = TRUE),
    n = n()) %>%
  # Filter to keep only stints with n => 10 per player
  dplyr::filter(!n < 9)
perform_stint_min
```

Let's take a quick look at the structure of our data now. 

```{r}
dplyr::glimpse(perform_stint_min)
```

# How many minutes should Giannis play per stint on court? 

Below, we create a very simple `ggplot` visualisation of the average number of rebounds effected per minute by Giannis Antetokounmpo for each period/stint that he was on the court over the 2020 season. A nice little *trick* that I have been using in recent months is to use `geom_smooth` instead of `geom_line` or `geom_point` to visualise the relationship between two continuous variables. This is by no means meant to be a predictive model - this is simply an exploratory data analysis (EDA) exercise. 

### (i) Number of rebounds 

```{r giannis rebounds per stint, warning = FALSE, message = FALSE, echo = TRUE}
perform_stint_min %>%
  dplyr::mutate(stints_on_court = as.factor(stints_on_court)) %>%
  dplyr::mutate(minute_cats = readr::parse_double(minute_cats)) %>%
  ggplot(data = ., aes(x = minute_cats, 
                       y = n_rebounds,
                       group = stints_on_court,
                       colour = stints_on_court)) +
  geom_smooth(aes(fill = stints_on_court),
              alpha = 0.3) +
   scale_y_continuous(limit=c(0,3),oob=squish) +
  #scale_y_continuous(limits = c(0, 2)) +
  labs(x = "Minutes on court (mins)",
       y = "No. of rebounds \n(per minute)",
       subtitle = "Each panel represents a period/stint that he was on the court") +
  facet_wrap(~ stints_on_court, ncol = 3)
```

Giannis seems to rebound really consistently. There is no apparent decline in rebounding rate the longer he is on the court.

### (ii) Number of 2PT FG's made per game 

```{r giannis 2pt fg per stint, warning = FALSE, message = FALSE, echo = TRUE}
perform_stint_min %>%
  #dplyr::mutate(stints_on_court = as.factor(stints_on_court)) %>%
  dplyr::mutate(minute_cats = readr::parse_double(minute_cats)) %>%
  dplyr::filter(stints_on_court < 6) %>%
  ggplot(data = ., aes(x = minute_cats, 
                       y = n_2pt_made,
                       group = stints_on_court,
                       colour = stints_on_court)) +
  geom_smooth(aes(fill = stints_on_court),
              alpha = 0.3) +
  labs(x = "Minutes on court (mins)",
       y = "No. of 2PT FG made \n(per minute)",
       subtitle = "Each panel represents a period/stint that he was on the court") +
  facet_wrap(~ stints_on_court, ncol = 3)
```

Giannis seems to make 2PT field goals quite consistently. There is no apparent decline in scoring rate the longer he is on the court. It is no surprise that Giannis is closing in on 11 000 points for the Bucks at only 26 years old.  

# How many minutes should Giannis play per game? 

Above we looked at Giannis's performance (no. rebounds and 2PT scoring) per minute player during successive stints on court. This data could help us to determine how long Giannis could be expected to perform at his peak capacity for each stint he has on court (hint: his performance doesn't decline over time). However, what if we wanted to see how many minutes Giannis should be playing per game to maximise performance? 

### (i) Number of rebounds 

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = "hide"}
# Calculate mean rebounds and 2PT's made per minute over the entire game 
perform_stint_game <- events_defined %>%
  dplyr::ungroup() %>%
  dplyr::filter(playerNames == "Giannis Antetokounmpo") %>%
  dplyr::filter(namePlayer1 == "Giannis Antetokounmpo") %>%
  dplyr::group_by(playerNames, idGame, stints_on_court, minute_cats, .drop = FALSE) %>%
  dplyr::summarise(
    total_rebounds = sum(is_rebound, na.rm = TRUE),
    total_2pt_made = sum(is_2pt_made, na.rm = TRUE),
    total_2pt_attempt = sum(!is.na(is_2pt_made)),
    perc_2pt_made = total_2pt_made / total_2pt_attempt,
    total_3pt_made = sum(is_3pt_made, na.rm = TRUE),
    total_3pt_attempt = sum(!is.na(is_3pt_made)),
    perc_3pt_made = total_3pt_made / total_3pt_attempt,
    total_ft_made  = sum(is_ft_made, na.rm = TRUE),
    total_ft_attempt = sum(!is.na(is_ft_made)),
    perc_ft_made = total_ft_made / total_ft_attempt) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(minute_cats = stringr::str_pad(minute_cats, 
                                               width = 3,
                                               side = "left", 
                                               pad = 0)) %>%
  dplyr::arrange(playerNames, idGame, stints_on_court, minute_cats) %>%
  dplyr::group_by(playerNames, idGame) %>%
  dplyr::mutate(total_minutes_game = row_number()) %>%
  dplyr::ungroup() %>%
  # Now calculate mean summary for season (not per-game)
  dplyr::group_by(playerNames, total_minutes_game, .drop = FALSE) %>%
  dplyr::summarise(
    n_rebounds       = mean(total_rebounds, na.rm = TRUE),
    n_2pt_made       = mean(total_2pt_made, na.rm = TRUE),
    n_3pt_made       = mean(total_3pt_made, na.rm = TRUE),
    n_ft_attempt     = mean(total_ft_made, na.rm = TRUE),
    sd_rebounds      = sd(total_rebounds, na.rm = TRUE),
    sd_2pt_made      = sd(total_2pt_made, na.rm = TRUE),
    sd_3pt_made      = sd(total_3pt_made, na.rm = TRUE),
    sd_ft_attempt    = sd(total_ft_made, na.rm = TRUE),
    n = n()) %>%
  # Filter to keep only stints with n => 10 per player
  dplyr::filter(!n < 10)
```

```{r giannis rebounds total minutes, warning = FALSE, message = FALSE, echo = TRUE}
perform_stint_game %>%
  ggplot(data = ., aes(x = total_minutes_game, 
                       y = n_rebounds)) +
  geom_smooth(colour = "black") +
  labs(x = "Total minutes played (mins)",
       y = "No. of rebounds \n(per minute)") 
```

Giannis gets plenty of rebounds throughout the game, but the rate at which he rebounds does decrease over the course of the game, but increases in the last 10 minutes. There is a small lull in rebounding (<0.1 fewer rebounds) after 10 - 20 minutes on court versus his first 10 minutes. All things considered, Giannis's rebounding rate is pretty consistent over an entire game.

### (ii) Number of 2PT FG's made per game 

```{r giannis 2pt total minutes, warning = FALSE, message = FALSE, echo = TRUE, preview = TRUE}
perform_stint_game %>%
  #dplyr::mutate(stints_on_court = as.factor(stints_on_court)) %>%
  #dplyr::mutate(total_m = readr::parse_double(minute_cats)) %>%
  dplyr::filter(playerNames == "Giannis Antetokounmpo") %>%
  ggplot(data = ., aes(x = total_minutes_game, 
                       y = n_2pt_made)) +
  geom_smooth(colour = "black") +
  labs(x = "Total minutes played (mins)",
       y = "No. of 2PT FG's made \n(per minute)") 
```

Giannis can clearly maintain his optimal 2PT scoring rate over the duration of an entire game.  

# What R skills did we learn today?

- Download NBA play-by-play data from the NBA API
- Process the data using common `tidyverse` cleaning packages (e.g. `dplyr`, `tidyr`, `stringr`)
- Create simple `ggplot` graphics to visualise player performance over time


#### Analytics implications:

The workflow presented above is how I approach exploratory data analysis (EDA) when building predictive models. I like to invest a lot of time into my EDA to get a really good understanding of the underlying data I am working with. In terms of model building, I like to only include predictor variables that appear to have some justifiable relationship with whatever our response variable is. Moreover, I like to have some idea of what form the function of y ~ x should take. For example, if we were trying to build a model to predict how many points Giannis will score per game, our EDA suggests that a linear relationship between minutes played per game and number of 2PT field goals *MAY* be appropriate. If we had found evidence that his 2PT scoring rate declines over time, we would then want to fit a non-linear trend for this relationship. There was a small dip in scoring rates in the 10 - 20 minute period, so we still may want a non-linear term in our model, but we could start with a linear term and build from there. 

#### Practical implications:

Giannis's performance is consistent over the course of a game. There is no obvious drop-off in his rebounding rate or 2PT scoring rates, both with respect to how many minutes he plays in each game, and with respect to how many stints on court he has and how long each stint is. It would be interesting to see how he performs above 35+ minutes per game... 






