---
title: "Do NBA players get tired?"
description: |
  NBA players are ATHLETES. But, do they get tired? Does their performance decline late in the game? 
author:
  - name: Guy Frederick Sutton
    url: {}
date: 01-25-2021
output:
  distill::distill_article:
    self_contained: false
draft: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

It always amazes me watching professional athletes at work. If I had to run, shoot, defend, sprint, turn like that... I would be lying on the floor, gasping for air within the 1st quarter. This got me thinking, do NBA players get tired? Does their performance vary over time? Is their performance as good in the 3rd or 4th quarter as in the 1st quarter? In my first blog post, I will be exploring this question using some basic data visualisations. Later blogposts will get into a bit more detail (building some statistical models) - stay tuned! 

# Load required packages 

The workhorse package we will be using is the NBA stats API package [`nbastatR`](https://github.com/abresler/nbastatR). We need to install the package from GitHub, which only needs to be done once. 

```{r install git package, include = TRUE}
# To install a GitHub package: 
devtools::install_github("abresler/nbastatR")

# Once you have downloaded a GitHub package, we should then
# edit our code so that the package is not re-installed each time we 
# re-run our script. We place a '#' in front of the command to comment it out. 
# If we need to run that code again later, we simply remove the '#' 
# devtools::install_github("abresler/nbastatR")
```

Our next step is to load the other packages we will require for this analysis. I really like using the `pacman` package to install/load R packages. I find this really helps when sharing code with collaborators who are less familiar with R. They can just open the script and press 'run' and not have to worry about installing new packages. `pacman` will install any packages that are not installed on your machine, and load all the remaining packages. 

```{r, include = TRUE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               lubridate,
               zoo,
               janitor,
               nbastatR,
               future)
```

# Downloading play-by-play data 

We will now use the `nbastatR` package to download play-by-play data from the NBA API. We can download data for different seasons, different teams... This can take a really long time (at least it does on my PC), so below, I will just download the data for (1) the 2019/2020 season, and (2) only the New Orleans Pelicans. 

Once the data is downloaded, quite a bit of data pre-processing and data cleaning is required, which will not be shown in this blogpost. I have edited this code slightly from the amazing [NBA blog by Ramiro Bentes](https://nbainrstats.netlify.app/post/adding-lineups-to-nba-play-by-play-data/). If you want to delve into the data cleaning code, it is available in the [source R markdown file] for this blogpost (https://github.com/guysutton/nba_analytics_blog/blob/main/_posts/2021-01-25-do-nba-players-get-tired/do-nba-players-get-tired.Rmd). 

```{r download game logs, include = TRUE}
# Extract all games for the 2020 season (excluding pre-season and all-star weekend)
game_logs <- game_logs(seasons = 2020)

# Process the game logs 
games <- game_logs %>%
  # Keep only these columns 
  dplyr::select(idGame, slugTeam, slugOpponent, locationGame) %>%
  # Use a conditional statement to process the team home/away data 
  dplyr::mutate(slugTeamHome = dplyr::if_else(locationGame == "H", 
                                              slugTeam, 
                                              slugOpponent),
                slugTeamAway = dplyr::if_else(locationGame == "A", 
                                              slugTeam, 
                                              slugOpponent)) %>%
  # Drop these columns - no longer required
  dplyr::select(-c(slugTeam, slugOpponent, locationGame)) %>%
  dplyr::distinct(idGame, .keep_all = TRUE) %>%
  # Here, we filter for the teams we want data for 
  # We have to use team codes. E.g. NOP = New Orleans Pelicans 
  # - Filter the games where either (denoted by the '|' bar) the home or 
  #   or away team is the Pelicans. 
  dplyr::filter(slugTeamHome == "NOP" | slugTeamAway == "NOP")
```

```{r background data cleaning}
############################################################
# - Pre-processing data cleaning
############################################################

# Lots, and lots, and lots of data cleaning 
new_pbp <- play_logs_all %>%
  dplyr::distinct(idGame, numberEvent, .keep_all = TRUE) %>%
  dplyr::group_by(idGame) %>%
  dplyr::mutate(numberEvent = row_number()) %>%  
  dplyr::ungroup() %>%
  dplyr::select(idGame, numberEventMessageType, numberEventActionType, 
                namePlayer1, namePlayer2, namePlayer3,  
                slugTeamPlayer1, slugTeamPlayer2,  slugTeamPlayer3, 
                numberPeriod, timeQuarter, minuteRemainingQuarter,  
                secondsRemainingQuarter, descriptionPlayHome, numberEvent, 
                descriptionPlayVisitor, scoreHome, scoreAway) %>%
  dplyr::mutate(shotPtsHome = dplyr::case_when(
    numberEventMessageType == 3 & !stringr::str_detect(descriptionPlayHome, "MISS") ~ 1,         
    numberEventMessageType == 1 & stringr::str_detect(descriptionPlayHome, "3PT") ~ 3,          
    numberEventMessageType == 1 & !stringr::str_detect(descriptionPlayHome, "3PT") ~ 2,
    TRUE ~ 0)) %>%
  dplyr::mutate(shotPtsAway = dplyr::case_when(
    numberEventMessageType == 3 & !stringr::str_detect(descriptionPlayVisitor, "MISS") ~ 1,
    numberEventMessageType == 1 & stringr::str_detect(descriptionPlayVisitor, "3PT") ~ 3,
    numberEventMessageType == 1 & !stringr::str_detect(descriptionPlayVisitor, "3PT") ~ 2,
    TRUE ~ 0)) %>%
  dplyr::group_by(idGame) %>%
  dplyr::mutate(ptsHome = cumsum(shotPtsHome),
                ptsAway = cumsum(shotPtsAway)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(secsLeftQuarter = (minuteRemainingQuarter * 60) + secondsRemainingQuarter) %>%
  dplyr::mutate(secsStartQuarter = dplyr::case_when(
    numberPeriod %in% c(1:5) ~ (numberPeriod - 1) * 720,
    TRUE ~ 2880 + (numberPeriod - 5) * 300)) %>%
  dplyr::mutate(secsPassedQuarter = ifelse(numberPeriod %in% c(1:4), 
                                           720 - secsLeftQuarter, 
                                           300 - secsLeftQuarter),
                secsPassedGame = secsPassedQuarter + secsStartQuarter) %>%
  dplyr::left_join(games %>%
                     dplyr::select(idGame, slugTeamHome, slugTeamAway)) %>%
  dplyr::select(idGame, numberEventMessageType, numberEventActionType, 
                slugTeamHome, slugTeamAway, slugTeamPlayer1, slugTeamPlayer2, 
                slugTeamPlayer3, numberPeriod, timeQuarter, secsPassedGame, numberEvent, 
                namePlayer1, namePlayer2, namePlayer3, descriptionPlayHome,
                descriptionPlayVisitor, ptsHome, ptsAway, shotPtsHome, shotPtsAway) %>%
  dplyr::mutate(marginBeforeHome = ptsHome - ptsAway - shotPtsHome + shotPtsAway,
                marginBeforeAway = ptsAway - ptsHome - shotPtsAway + shotPtsHome,
                timeQuarter = stringr::str_pad(timeQuarter, width = 5, pad = 0))

# Define where substitutes where made 
subs_made <- new_pbp %>%
  dplyr::filter(numberEventMessageType == 8) %>%
  dplyr::mutate(slugTeamLocation = ifelse(slugTeamPlayer1 == slugTeamHome, "Home", "Away")) %>%
  dplyr::select(idGame, numberPeriod, timeQuarter, secsPassedGame, 
                slugTeamPlayer = slugTeamPlayer1,
                slugTeamLocation, 
                playerOut = namePlayer1, 
                playerIn = namePlayer2) %>%
  tidyr::pivot_longer(cols = starts_with("player"), 
                      names_to = "inOut",
                      names_prefix = "player",
                      values_to = "namePlayer") %>%
  dplyr::group_by(idGame, numberPeriod, slugTeamPlayer, namePlayer) %>%
  dplyr::filter(row_number() == 1) %>%
  dplyr::ungroup()

others_qtr <- new_pbp %>%
  dplyr::filter(numberEventMessageType != 8) %>%                             
  dplyr::filter(!(numberEventMessageType == 6 & 
                    numberEventActionType %in% c(10, 11, 16, 18, 25))) %>%
  tidyr::pivot_longer(cols = starts_with("namePlayer"),
                      names_to = "playerNumber",
                      names_prefix = "namePlayer",
                      values_to = "namePlayer") %>%
  dplyr::mutate(slugTeamPlayer = dplyr::case_when(playerNumber == 1 ~ slugTeamPlayer1,
                                                  playerNumber == 2 ~ slugTeamPlayer2,
                                                  playerNumber == 3 ~ slugTeamPlayer3,
                                                  TRUE ~ "None")) %>%
  dplyr::mutate(slugTeamLocation = ifelse(slugTeamPlayer == slugTeamHome, 
                                          "Home", 
                                          "Away")) %>%
  dplyr::filter(!is.na(namePlayer),
                !is.na(slugTeamPlayer)) %>%
  dplyr::anti_join(subs_made %>%
                     dplyr::select(idGame, numberPeriod, slugTeamPlayer, namePlayer)) %>%  
  dplyr::distinct(idGame, numberPeriod, namePlayer, slugTeamPlayer, slugTeamLocation)

lineups_quarters <- subs_made %>%
  dplyr::filter(inOut == "Out") %>%
  dplyr::select(idGame, numberPeriod, slugTeamPlayer, namePlayer, slugTeamLocation) %>%
  dplyr::bind_rows(others_qtr) %>%
  dplyr::arrange(idGame, numberPeriod, slugTeamPlayer)

lineup_subs <- new_pbp %>%
  dplyr::filter(numberEventMessageType == 8) %>%
  dplyr::select(idGame, numberPeriod, timeQuarter, secsPassedGame, 
                slugTeamPlayer = slugTeamPlayer1, 
                playerOut = namePlayer1, 
         playerIn = namePlayer2, numberEvent) %>%
  dplyr::arrange(idGame, numberEvent) %>%
  dplyr::group_by(idGame, numberPeriod, slugTeamPlayer) %>%
  dplyr::mutate(row1 = row_number()) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(lineups_quarters %>%
                     dplyr::group_by(idGame, numberPeriod, slugTeamPlayer) %>%
                     dplyr::summarise(lineupBefore = paste(sort(unique(namePlayer)), 
                                                           collapse = ", ")) %>%
                     dplyr::ungroup() %>%
                     dplyr::mutate(row1 = 1)) %>%
  dplyr::select(-row1)

lineup_subs <- lineup_subs %>%
  dplyr::mutate(lineupBefore = stringr::str_split(lineupBefore, ", ")) %>% 
  dplyr::arrange(idGame, numberEvent) %>%
  dplyr::group_by(idGame, numberPeriod, slugTeamPlayer) %>%
  dplyr::mutate(lineupAfter = accumulate2(playerIn, 
                                          playerOut, 
                                          ~setdiff(c(..1, ..2), ..3), 
                                          .init = lineupBefore[[1]])[-1],
         lineupBefore = dplyr::coalesce(lineupBefore, lag(lineupAfter))) %>%
  dplyr::ungroup() %>% 
  dplyr::mutate_all( ~ purrr::map_chr(., ~paste(.x, collapse = ", "))) %>%
  dplyr::mutate_at(vars("numberEvent", "numberPeriod", "idGame"), 
                   ~ as.integer(.)) %>%
  dplyr::mutate(secsPassedGame = as.numeric(secsPassedGame)) %>%
  dplyr::arrange(idGame, numberEvent) %>%
  dplyr::left_join(lineups_quarters %>%
                     dplyr::distinct(idGame, slugTeamPlayer, slugTeamLocation)) %>%
  dplyr::filter(!is.na(slugTeamLocation))

lineup_game <- new_pbp %>%
  dplyr::group_by(idGame, numberPeriod) %>%
  dplyr::mutate(row1 = row_number()) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(lineups_quarters %>%
                     dplyr::group_by(idGame, numberPeriod, slugTeamLocation) %>%
                     dplyr::summarise(lineupBefore = paste(sort(unique(namePlayer)), 
                                                           collapse = ", ")) %>%
                     dplyr::ungroup() %>%
                     tidyr::pivot_wider(names_from = slugTeamLocation,
                                        names_prefix = "lineupInitial",
                                        values_from = lineupBefore) %>%
                     dplyr::mutate(row1 = 1)) %>%
  dplyr::select(-row1) %>%
  dplyr::left_join(lineup_subs %>%
              mutate(lineupBeforeHome = ifelse(slugTeamLocation == "Home", lineupBefore, NA),
                     lineupAfterHome = ifelse(slugTeamLocation == "Home", lineupAfter, NA),
                     lineupBeforeAway = ifelse(slugTeamLocation == "Away", lineupBefore, NA),
                     lineupAfterAway = ifelse(slugTeamLocation == "Away", lineupAfter, NA)) %>%
              select(idGame, numberPeriod, timeQuarter, secsPassedGame, 
                     numberEvent, slugTeamPlayer1 = slugTeamPlayer,
                     contains("Home"), 
                     contains("Away"))) %>%
  dplyr::mutate_at(vars(c(lineupBeforeHome, lineupAfterHome)), 
                   ~ ifelse(!is.na(lineupInitialHome), 
                            lineupInitialHome, 
                            .)) %>%
  dplyr::mutate_at(vars(c(lineupBeforeAway, lineupAfterAway)), 
                   ~ ifelse(!is.na(lineupInitialAway), 
                            lineupInitialAway, 
                            .)) %>%
  dplyr::select(-starts_with("lineupInitial"))

lineup_game <- lineup_game %>%
  dplyr::group_by(idGame, numberPeriod) %>%
  dplyr::mutate(lineupHome = na.locf(lineupAfterHome, na.rm = FALSE),
                lineupAway = na.locf(lineupAfterAway, na.rm = FALSE),
                lineupHome = ifelse(is.na(lineupHome), 
                                    na.locf(lineupBeforeHome, 
                                            fromLast = TRUE, 
                                            na.rm = FALSE), 
                                    lineupHome),
         lineupAway = ifelse(is.na(lineupAway), 
                             na.locf(lineupBeforeAway, 
                                     fromLast = TRUE, 
                                     na.rm = FALSE), 
                             lineupAway),
         lineupHome = stringr::str_split(lineupHome, ", "),
         lineupAway = stringr::str_split(lineupAway, ", "),
         lineupHome = purrr::map_chr(lineupHome, 
                                     ~ paste(sort(.), 
                                             collapse = ", ")),
         lineupAway = purrr::map_chr(lineupAway, 
                                     ~ paste(sort(.), 
                                             collapse = ", "))) %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(starts_with("lineupBefore"), 
                   starts_with("lineupAfter")))

lineup_game_stats <- lineup_game %>%
  dplyr::mutate(canSub = case_when(numberEventMessageType == 5 & 
                                     !numberEventActionType %in% c(1, 2) ~ 1,
                                   numberEventMessageType == 6 & 
                                     numberEventActionType != 16 ~ 1,
                                   numberEventMessageType == 11 & 
                                     numberEventActionType != 4 ~ 1,
                                   numberEventMessageType == 7 & 
                                     numberEventActionType == 5 ~ 1, 
                                   numberEventMessageType == 4 & 
                                     numberEventActionType == 0 & 
                                     !stringr::str_detect(stringr::str_to_upper(descriptionPlayHome),
                                                   "OFF:") ~ 1,
                                   numberEventMessageType == 4 & 
                                     numberEventActionType == 0 & 
                                     !str_detect(stringr::str_to_upper(descriptionPlayVisitor), 
                                                 "OFF:") ~ 1,
                                   TRUE ~ 0)) %>%
  dplyr::mutate(secsPassedGame2 = ifelse(timeQuarter == "12:00" &
                                           (str_detect(str_to_lower(descriptionPlayHome), 
                                                       "technical") |
                                              str_detect(str_to_lower(descriptionPlayVisitor),
                                                         "technical")), 
                                         secsPassedGame + 0.5, secsPassedGame)) %>% 
  dplyr::group_by(idGame, numberPeriod, secsPassedGame) %>%
  dplyr::mutate(numberNew = ifelse(numberEventMessageType == 3 & 
                                     numberEventActionType == 12, 
                                   paste(numberEvent[numberEventMessageType == 3 &
                                                       numberEventActionType == 11], 
                                         collapse = ", "), 
                                   as.character(numberEvent)),
                numberNew = ifelse(numberEventMessageType == 3 & 
                                     numberEventActionType %in% c(14, 15), 
                                   paste(numberEvent[numberEventMessageType == 3 & 
                                                       numberEventActionType == 13], 
                                         collapse = ", "),
                                   numberNew)) %>%
  dplyr::mutate(numberNew = stringr::str_split(numberNew, ", "),
                numberNew = purrr::map(numberNew, ~as.numeric(.)),
                numberNew = purrr::map2_dbl(numberNew, 
                                            numberEvent, 
                                            ~ max(.x[.x <= .y]))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(idGame, numberNew, numberEvent) %>%
  dplyr::group_by(idGame) %>%
  dplyr::mutate(newptsHome = cumsum(shotPtsHome),
                newptsAway = cumsum(shotPtsAway)) %>%
  dplyr::group_by(idGame, numberPeriod, secsPassedGame2) %>%
  dplyr::mutate(subOpp = cumsum(canSub)) %>%
  dplyr::group_by(idGame = as.character(idGame),
                  numberPeriod = as.character(numberPeriod), 
                  subOpp, 
                  secsPassedGame2 = as.character(secsPassedGame2)) %>%
  dplyr::mutate(hasFouls = sum(numberEventMessageType == 3)) %>%
  dplyr::mutate(newptsHome = ifelse(hasFouls > 0,
                                    newptsHome[row_number() == 
                                                 max(row_number()[numberEventMessageType == 3])],
                                    newptsHome),
                newptsAway = ifelse(hasFouls > 0,
                                    newptsAway[row_number() == 
                                                 max(row_number()[numberEventMessageType == 3])],
                                    newptsAway)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-hasFouls) %>%
  dplyr::select(-c(numberNew, secsPassedGame2)) %>%
  dplyr::mutate_all(~ as.character(.)) %>%
  dplyr::mutate(secsPassedGame = as.numeric(secsPassedGame),
                numberEvent = as.numeric(numberEvent))
```

We now have play-by-play data for all the New Orleans Pelicans games for the 2019/2020 season. Each game has a unique ID number (`idGame`), each play within each game has its own unique ID (`numberEvent`), and for each play, we have the players who were on the court for each team (`homeTeamLineup` and `awayTeamLineup`). 

```{r}
head(games)
```

However, we still need to process the data so that we can:

- Define how many periods each player was on the court for during each game
- Define how long was each period 
- Convert our data into long-format, so that each player who was on the court, for each play, gets their own row. (We will need this format of data when we start summarising how player performance later on). 

```{r}
# For each event, create a distinct row for each player from each team 
game_test <- lineup_game_stats %>%
  # Drop the events that are not plays (e.g. technical fouls)
  tidyr::drop_na(namePlayer1) %>%
  # Give a unique identified to each play per game 
  dplyr::group_by(idGame, numberEvent) %>%
  dplyr::mutate(numberEvent = dplyr::cur_group_id()) %>%
  # Duplicate the full team line-up. We need it for later. 
  dplyr::mutate(homeTeamLineup = lineupHome,
                awayTeamLineup = lineupAway) %>%
  # Combine team line-ups into a single column
  tidyr::unite(homeTeamLineup:awayTeamLineup, col = "lineups", sep = ", ") %>%
  tidyr::separate_rows(., 
                       lineupHome, 
                       lineupAway,
                       sep = ",", 
                       convert = TRUE) %>%
  # Remove whitespace from start of player names, if present
  dplyr::mutate(lineupHome = stringr::str_trim(lineupHome, side = c("left")),
                lineupAway = stringr::str_trim(lineupAway, side = c("left"))) %>%
  # Convert players names into long-format so each player now has their own row 
  # for each event in the game
  tidyr::pivot_longer(
    cols = lineupHome:lineupAway,
    names_to = "teamHomeAway",
    values_to = "playerNames") %>%
  dplyr::ungroup() %>%
  # For each event, is the lineup the same as the previous event? 
  dplyr::group_by(idGame) %>%
  dplyr::mutate(lineupChange = lineups != lag(lineups),
                lineupChange = coalesce(lineupChange, FALSE)) %>%
  dplyr::mutate(lineupStint = cumsum(lineupChange)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(idGame, lineupStint, numberEvent) %>%
  # Has that player been subbed? 
  dplyr::group_by(idGame, playerNames) %>%
  dplyr::arrange(idGame, playerNames, numberEvent) %>%
  # Add column which states whether player was subbed or not? 
  dplyr::mutate(player_subbed = dplyr::case_when(
    lag(numberEvent) != numberEvent - 1 ~ "Yes", 
    TRUE ~ "No"))
```

We now know exactly which plays each player was on the court for, but we need to create a column which explicitly notes whether the player was on the court or not (`stints`), and how long each stint on court lasted. 

```{r}
# Define each stint each player has on court 
define_stint <- game_test %>%
  # Define stints as the cumulative sum of times the player is subbed
  # cumsum starts at 0, so we need to + 1 to each value 
  dplyr::mutate(stints_on_court = cumsum(player_subbed == "Yes"),
                stints_on_court = stints_on_court + 1) %>%
  dplyr::ungroup() %>%
  # Define start and end time of each stint with a min/max statement
  dplyr::group_by(idGame, playerNames, stints_on_court) %>%
  dplyr::mutate(time_at_start = min(secsPassedGame),
                time_at_end   = max(secsPassedGame)) %>%
  # Calculate total seconds of each stint
  dplyr::mutate(time_stint_secs = time_at_end - time_at_start) %>%
  # Convert seconds into minutes and seconds 
  dplyr::mutate(time_stint_mins = time_stint_secs / 60) %>%
  # Calculate time since start of each stint 
  dplyr::mutate(time_since_start_stint = secsPassedGame - time_at_start) %>%
  # Manually define 60 second intervals - 
  # I couldn't force ggplot2::cut_width to start @ 0... Help?  
  dplyr::mutate(minute_cats = dplyr::case_when(
    between(time_since_start_stint, 0, 60) ~ "1",
    between(time_since_start_stint, 61, 120) ~ "2",
    between(time_since_start_stint, 121, 180) ~ "3",
    between(time_since_start_stint, 181, 240) ~ "4",
    between(time_since_start_stint, 241, 300) ~ "5",
    between(time_since_start_stint, 301, 360) ~ "6",
    between(time_since_start_stint, 361, 420) ~ "7",
    between(time_since_start_stint, 421, 480) ~ "8",
    between(time_since_start_stint, 481, 540) ~ "9",
    between(time_since_start_stint, 541, 600) ~ "10",
    between(time_since_start_stint, 601, 660) ~ "11",
    between(time_since_start_stint, 661, 720) ~ "12",
    between(time_since_start_stint, 721, 780) ~ "13",
    between(time_since_start_stint, 781, 840) ~ "14",
    between(time_since_start_stint, 841, 900) ~ "15",
    between(time_since_start_stint, 901, 960) ~ "16",
    between(time_since_start_stint, 961, 1020) ~ "17",
    between(time_since_start_stint, 1021, 1080) ~ "18",
    between(time_since_start_stint, 1081, 1140) ~ "19",
    between(time_since_start_stint, 1141, 1200) ~ "20",
    between(time_since_start_stint, 1221, 1280) ~ "21",
    between(time_since_start_stint, 1281, 1340) ~ "22",
    between(time_since_start_stint, 1341, 1400) ~ "23",
    between(time_since_start_stint, 1401, 1460) ~ "24")) %>%
  # For each play, does the score change or not?
  # - We will need this to classify whether 2pt/3pt/free throw attempts
  #   were successful or not. 
  dplyr::mutate(score_change = dplyr::if_else(shotPtsHome > 0 | 
                                                shotPtsAway > 0, 1, 0))
```

Our last bit of data cleaning requires that we categorize each play. For example, was the play a 2pt field goal attempt, a free throw attempt, was the shot made or missed? 

```{r}
# Combine the descriptions for each play for home and away teams into 
# a single column
description_play <- minute_groups %>%
  dplyr::mutate(descriptionPlay = dplyr::coalesce(descriptionPlayHome, descriptionPlayVisitor))

# Define each event as a particular play
events_defined <- description_play %>%
  # Was 3pt fg attempt made?
  dplyr::mutate(is_3pt_made = dplyr::case_when(
    stringr::str_detect(descriptionPlay, "3PT") &
      score_change == 1 ~ 1,
    stringr::str_detect(descriptionPlay, "3PT") &
      score_change == 0 ~ 0)) %>%
  # Was 2pt fg attempt made?
  dplyr::mutate(is_2pt_made = dplyr::case_when(
    numberEventMessageType == 1 & 
      !stringr::str_detect(descriptionPlay, "3PT") ~ 1,
    numberEventMessageType == 2 & 
      !stringr::str_detect(descriptionPlay, "3PT") ~ 0)) %>%
  # Was event a rebound? 
  dplyr::mutate(is_rebound = dplyr::case_when(
    numberEventMessageType == 4 & 
      stringr::str_detect(descriptionPlay, "REBOUND") ~ 1,
    TRUE ~ 0)) %>%
  # Was event a made free throw?
  dplyr::mutate(is_ft_made = dplyr::case_when(
    numberEventMessageType == 3 & 
      score_change == 1 ~ 1,
    numberEventMessageType == 3 & 
      score_change == 0 ~ 0))
```

# Summarise and visualise performance over time 

Finally, the bit of the analysis that we are actually most interested in. Let's now summarise how many 2pt/3pt/free throws/rebounds are made per minute that each player is on the court, for each game, and then these summarise performance metrics over the entire 2020 season. 

```{r}
perform_stint_min <- events_defined %>%
  dplyr::ungroup() %>%
  dplyr::group_by(idGame, playerNames, stints_on_court, minute_cats, .drop = FALSE) %>%
  dplyr::summarise(
    total_rebounds = sum(is_rebound, na.rm = TRUE),
    total_2pt_made = sum(is_2pt_made, na.rm = TRUE),
    total_2pt_attempt = sum(!is.na(is_2pt_made)),
    perc_2pt_made = total_2pt_made / total_2pt_attempt,
    total_3pt_made = sum(is_3pt_made, na.rm = TRUE),
    total_3pt_attempt = sum(!is.na(is_3pt_made)),
    perc_3pt_made = total_3pt_made / total_3pt_attempt,
    total_ft_made  = sum(is_ft_made, na.rm = TRUE),
    total_ft_attempt = sum(!is.na(is_ft_made)),
    perc_ft_made = total_ft_made / total_ft_attempt) %>%
  dplyr::ungroup() %>%
  # Now calculate mean summary for season (not per-game)
  dplyr::group_by(playerNames, stints_on_court, minute_cats, .drop = FALSE) %>%
  dplyr::summarise(
    total_rebounds = mean(total_rebounds, na.rm = TRUE),
    total_2pt_made = mean(total_2pt_made, na.rm = TRUE),
    total_3pt_made = mean(total_2pt_made, na.rm = TRUE),
    total_ft_attempt = mean(total_ft_made, na.rm = TRUE))
head(perform_stint_min)
```


```{r}
perform_stint_min %>%
  dplyr::mutate(stints_on_court = as.factor(stints_on_court)) %>%
  dplyr::mutate(minute_cats = readr::parse_double(minute_cats)) %>%
  dplyr::filter(playerNames == "Brandon Ingram") %>%
  ggplot(data = ., aes(x = minute_cats, 
                       y = total_2pt_made,
                       group = stints_on_court,
                       colour = stints_on_court)) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ stints_on_court, ncol = 5)

```







